---
title: Improving the candires MIC models
author: Joe Lewis
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    mathspec: true
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

## Building on the random effects model

The current MIC models I've built have a random intercept per participant to
account for repeated sampling of people. This assumes lognormally distributed
MICs and fits a different mean for the distribution for each participant. This
is necessary because the model needs to account for within-person correlation of
MIC; but it's pretty crude and probably doesn't represent what we want. The
effect is fixed throughout time, but the true individual effect probably decays
with time. Also you can imagine that if a participant has a low MIC at time $a$
and high MIC at time $b$ the model will fit a baseline risk between the two,
which might not be what we want, because this *change* is what we are interested
in. It might pull some of the drug effect into the random intercept. 

An possible improvement on this could be a *Gaussian process* model. This
assumes that samples within participants are correlated in time, with a decaying
effect the time scale of which is fitted by the model. Given enough time
separation, the within-participant correlation falls to 0. This is probably a
better representation of our conceptual model of what is happening. 

Formally, we have $n$ measurements of a normally distributed outcome variable (i.e.
log-transformed MIC) for $N$ participants; each may have a different number of
measurements. We can construct a regression model where the measurements $y_{i},
i = 1,2 ... n$ are given by:

$$
y_{i} \sim \text{Normal}(\mu_{i}, \sigma_{MIC}) 
$$

I.e. each sample is drawn from a normal distribution with mean $\mu_{i}$ but a
fixed population standard deviation $\sigma_{MIC}$.

Switching to vector notation (bold type lowercase indicates a vector, bold
uppercase a matrix) we can define the $i$ length vector $\boldsymbol{\mu}$ as

$$
\boldsymbol{\mu} = \mu_{0} + \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{f}
$$

Where the scalar $\mu_{0}$ is the intercept, $\boldsymbol{\beta}$ is the vector
of covariates (length $c$ equal to number of covariates), $\boldsymbol{X}$ is
the design matrix of covariate values (with $c$ columns and $n$ rows, one per
sample), and $\boldsymbol{f}$ is a length $n$ vector with the gaussian process
effect per observation. This is drawn from a multivariate normal distribution: 

$$
\boldsymbol{f} \sim \text{MVN}(0, \boldsymbol{\Sigma})
$$

Where $\boldsymbol{\Sigma}$ is the covariance matrix; this encodes the temporal
correlation of the samples. We make it a block-diagonal covariance matrix equal
to 0 for samples from different participants, and an exponentiated quadratic
kernel for samples from the same participant. That is, for a given pair of
observations $i$ and $j$ with time difference $t_{ij}$ between them, the
covariance is

$$
\Sigma_{ij} = \alpha^{2} \text{exp}(-\frac{-t^{2}_{ij}}{2l^{2}}) 
$$


Here $\alpha$ encodes the magnitude of covariance and $l$ the temporal
correlation.  This is one of many possible choices: it's a common choice in
gaussian process models.

To make it easier computationally, and following the Stan user manual, we can
use the fact that it is possible to rescale a multivariate normal distribution
as the product of the Cholesky decomposition of the covariance matrix and an
isotropic normal variable, i.e:

$$
\boldsymbol{L\eta} \sim \text{MVN}(0,\boldsymbol{\Sigma})
$$
Where 
$$
\boldsymbol{\eta} \sim \text{Normal}(\boldsymbol{0}, \boldsymbol{I})  
$$
and  
$$
\boldsymbol{L}^{T}\boldsymbol{L} = \boldsymbol{\Sigma}
$$

i.e. $\boldsymbol{L}$ is the Cholesky decomposition of $\boldsymbol{\Sigma}$.

We define the covariate effect in exactly the same was as the simpler random
intercept models; that is, for a given sample $i$ and covariate $j$ the matrix
$X_{ij}$ is defined by:

$$
X_{ij} = 
  \begin{cases}
  0 & \text{if not antifungal exposure for covariate $j$}\\
  1 & \text{if ongoing antifungal exposure for covariate $j$}\\
  e^{\frac{-t_{ij}}{\tau}} & \text{if prior antifungal exposure a time $t_{ij}$
    ago for covariate $j$}\\
  \end{cases}
$$


## Back to a simpler approach

Taking a step back, this modelling approach assesses for a population shift in
MIC following antifungal exposure. But an alternate conception of this is that
there is a stochastic aquisition of resistance associated mutations, which will
result in no change in MIC for most samples, but some will develop resistance.
So let's try a dichotomised model - define "acquisition of resistance" as a
log~2~ fold increase of 2 or more, and plot the proportion of participants who
have this change from baseline. Because the sampling is all at different times,
we can use a LOESS regression line to estimate the proportion of participants
who have acquisition of resistance, as defined in this way, and plot id against
days since drug exposure. We need to compare to the antifungal unexposed group -
to put them on the same axis I've aribtrarily defined "days since drug exposure"
to be "days since enrollment." I've also tabulated the number of isolates.

To do some statistical tests I've followed Phoebe's original analysis -
comparing people without a given drug exposure to those with an exposure and
looking at proportion of those with a log~2~ fold increase in MIC of 2 or more.
I've then used a Fisher test to test for equality of these proportions.

# Results

## Log~2~ change in MIC

Start with the simple approach. The log2 fold change in MIC compared baseline
MIC is shown in the plots below, as well as an estimated proportion of samples
that have a greater than or equal to log~2~ fold change in MIC compared to
baseline. The actual numbers of isolates are shown in the tables, as are the
proportion of participants with emergent resistance.

The first thing to say is that there are small numbers of parapsilosis and
dublinensis which make drawing any conclusion difficult, and that a log~2~ fold
increase in MIC of 2 or more happens infrequently across all species for both
azole and echionocandin. However, when it does happen, it is associated with
drug exposure - this is most clearly the case in fluconazole and azole exposure
in albicans (p = 0.006) and glabrata (p = 0.02). Wide confidence intervals make
it more difficult to interpret anidulafungin MIC and echinocandin exposure but
there were more particioants with log~2~ (fold MIC increase) of at least 2 in
the albicans echincocandid exposed (6/54 [11%]) versus unexposed (3/102 [3%], p
= 0.07) and the glabrata echionocandin exposed (1/27 [4%]) versus unexposed
(0/35 [0%], p = 0.4)



```{r setup} 
#| include: false

library(tidyverse)
library(msm)
library(here)
library(ivs)
library(bayesplot)
library(patchwork)
library(kableExtra)


source(here("load_mic_data.R"))
source(here("modelling_helper_functions.R"))

library(cmdstanr)
```

```{r mic_fold_change_plots}

species <- c("albicans", "glabrata", "dubliniensis", "parapsilosis")
drug <- c("Azole", "Echinocandin")

listout <- list()
p <- 1
for (current_species in species) {
  for (current_drug in drug) {
    if (current_drug == "Azole") {
      current_mic_drug <- "fluconazole"
    } else {
      current_mic_drug <- "anidulafungin"
    }

    df_fold_change <-
      generate_mic_model_data(
        df_mics,
        df_drug,
        spec = current_species,
        exposure_drug_class = current_drug,
        mic_drug = current_mic_drug,
        code_antifungal_exposure_as = "zero_at_first_exposure"
      ) |>
      group_by(psn) |>
      arrange(psn, swb_date) |>
      mutate(
        baseline_mic = first(mic)
      ) |>
      ungroup() |>
      mutate(mic_fold_change = mic / baseline_mic)

    p_b <-
      df_fold_change |>
      ggplot(aes(days_since_first_drug_exposure,
        as.numeric(log2(mic_fold_change) >= 2),
        group = any_drug_exposure,
        color = any_drug_exposure
      )) +
      geom_smooth() +
      theme_bw() +
      theme(legend.position = "top") +
      labs(
        title = paste0(
          "Proportion of ",
          current_species,
          " >= 2 log2 fold change in\n",
          current_mic_drug,
          " MIC"
        ),
        subtitle = paste0(
          "Following ",
          current_drug,
          " exposure"
        ),
        y = "proportion"
      )

    p_a <-
      df_fold_change |>
      ggplot(aes(days_since_first_drug_exposure,
        log2(mic_fold_change),
        group = psn,
        color = any_drug_exposure
      )) +
      geom_point(alpha = 0.5) +
      geom_line() +
      theme_bw() +
      theme(legend.position = "top") +
      labs(
        title = paste0(
          current_species, " log2 fold change in ",
          current_mic_drug, " MIC"
        ),
        subtitle = paste0(
          "Following ", current_drug, " exposure"
        ),
        y = "log2(mic fold change)"
      )


    listout[[p]] <-
      p_a + p_b +
        plot_layout(guides = "collect") &
        theme(legend.position = "bottom")
    p <- p + 1
  }
}
```


```{r}
#| output: asis
res <- map2_chr(listout, list(1,2,3,4,5,6,7,8), \(plots, no) {
  knitr::knit_child(text = c(
    "```{r}",
    paste0("#|", " label: fig-micplot", no),
    '#| fig.cap: "log2[MIC fold change] (left) from baseline and proportion of isolates with 2 or higher log2 fold change (right). Lines link isolates from one participant in the left panel. Colors show if participants are antifungal exposed or unexposed."',
    "#| fig.height: 5",
    "#| fig.width: 10",
    "#| echo: false",
    "plots",
    "```",
    ""
  ), envir = environment(), quiet = TRUE)
})
cat(res, sep = "\n")
```

```{r} 
#| label: tbl-acqres1
#| tbl-cap: "Number of isolates that have a log2 fluconazole MIC fold change of at least 2, stratified by whether the patient has any azole exposure at all, and time since antimicrobial exposure. For those with no azole exposure at all, the days since drug exposure variable is set to days since enrolment"


current_drug <- drug[[1]]
if (current_drug == "Azole") {
  current_mic_drug <- "fluconazole"
} else {
  current_mic_drug <- "anidulafungin"
}



listout <- list()
p <- 1
for (current_species in species) {
  listout[[p]] <-
    generate_mic_model_data(
      df_mics,
      df_drug,
      spec = current_species,
      exposure_drug_class = current_drug,
      mic_drug = current_mic_drug,
      code_antifungal_exposure_as = "zero_at_first_exposure"
    ) |>
    group_by(psn) |>
    arrange(psn, swb_date) |>
    mutate(
      baseline_mic = first(mic)
    ) |>
    ungroup() |>
    mutate(
      mic_fold_change = mic / baseline_mic,
      days_since_exposure_cat = cut(
        days_since_first_drug_exposure,
        seq(from = -70, to = 1400, by = 14)
      ),
      acq_res = log2(mic_fold_change) >= 2
    ) |>
    mutate(
      acq_res = as.factor(acq_res),
      any_drug_exposure = as.factor(any_drug_exposure)
    ) |>
    count(days_since_exposure_cat, any_drug_exposure, acq_res, .drop = FALSE) |>
    group_by(days_since_exposure_cat, any_drug_exposure) |>
    mutate(total = sum(n)) |>
    filter(acq_res == "TRUE", total > 0) |>
    rowwise() |>
    mutate(
      prop = n / total,
      lci = binom.test(n, total)$conf.int[[1]],
      uci = binom.test(n, total)$conf.int[[2]]
    ) |>
    mutate(str = paste0(
      n, "/", total, " (", round(prop * 100, 0), "% [95% CI ",
      round(lci * 100, 0), "-", round(uci * 100, 0), "%])"
    )) |>
    select(c(days_since_exposure_cat, any_drug_exposure, str)) |>
    pivot_wider(
      id_cols = days_since_exposure_cat,
      names_from = any_drug_exposure,
      values_from = str,
      values_fill = "0/0"
    ) |>
    rename(drug_exposed = `TRUE`, drug_unexposed = `FALSE`) |>
  select(days_since_exposure_cat, drug_unexposed, drug_exposed)
  p <- p + 1
}



listout[[1]] |>
  left_join(listout[[2]],
    by = "days_since_exposure_cat"
  ) |>
  mutate(across(everything(), \(x) if_else(is.na(x), "0/0", x))) |>
  kbl(
    booktabs = TRUE,
    col.names =
      c(
        "days since\ndrug exposure",
        rep(c("Azole unexposed", "Azole exposed"), 2)
      )
  ) |>
  add_header_above(c(
    " " = 1, "albicans" = 2, "glabrata" = 2
  )) |>
  kable_styling(latex_options = "scale_down")
```

```{r} 
#| label: tbl-acqres2
#| tbl-cap: "Number of isolates that have a log2 fluconazole MIC fold change of at least 2, stratified by whether the patient has any azole exposure at all, and time since azole exposure. For those with no antimicrobial exposure at all, the days since drug exposure variable is set to days since enrolment"

listout[[2]] |>
  left_join(listout[[3]],
    by = "days_since_exposure_cat"
  ) |>
  mutate(across(everything(), \(x) if_else(is.na(x), "0/0", x))) |>
  kbl(
    booktabs = TRUE,
    col.names =
      c(
        "days since\ndrug exposure",
        rep(c("Azole unexposed", "Azole exposed"), 2)
      )
  ) |>
  add_header_above(c(
    " " = 1, "dubliniensis" = 2, "papapsilosis" = 2
  )) |>
  kable_styling(latex_options = "scale_down")
```




```{r} 
#| label: tbl-acqres3
#| tbl-cap: "Number of isolates that have a log2 anidulafungin MIC fold change of at least 2, stratified by whether the patient has any echinocandin exposure at all, and time since echinocandin exposure. For those with no echinocandin exposure at all, the days since drug exposure variable is set to days since enrolment"



species <- c("albicans", "glabrata", "dubliniensis", "parapsilosis")
current_drug <- drug[[2]]
if (current_drug == "Azole") {
  current_mic_drug <- "fluconazole"
} else {
  current_mic_drug <- "anidulafungin"
}
listout <- list()
p <- 1
for (current_species in species) {
  listout[[p]] <-
    generate_mic_model_data(
      df_mics,
      df_drug,
      spec = current_species,
      exposure_drug_class = current_drug,
      mic_drug = current_mic_drug,
      code_antifungal_exposure_as = "zero_at_first_exposure"
    ) |>
    group_by(psn) |>
    arrange(psn, swb_date) |>
    mutate(
      baseline_mic = first(mic)
    ) |>
    ungroup() |>
    mutate(
      mic_fold_change = mic / baseline_mic,
      days_since_exposure_cat = cut(
        days_since_first_drug_exposure,
        seq(from = -70, to = 1400, by = 14)
      ),
      acq_res = log2(mic_fold_change) >= 2
    ) |>
    mutate(
      acq_res = factor(as.character(acq_res), levels = c("FALSE", "TRUE")),
      any_drug_exposure = as.factor(any_drug_exposure)
    ) |>
    count(days_since_exposure_cat, any_drug_exposure, acq_res, .drop = FALSE) |>
    group_by(days_since_exposure_cat, any_drug_exposure) |>
    mutate(total = sum(n)) |>
    filter(acq_res == "TRUE", total > 0) |>
    rowwise() |>
    mutate(
      prop = n / total,
      lci = binom.test(n, total)$conf.int[[1]],
      uci = binom.test(n, total)$conf.int[[2]]
    ) |>
    mutate(str = paste0(
      n, "/", total, " (", round(prop * 100, 0), "% [95% CI ",
      round(lci * 100, 0), "-", round(uci * 100, 0), "%])"
    )) |>
    select(c(days_since_exposure_cat, any_drug_exposure, str)) |>
    pivot_wider(
      id_cols = days_since_exposure_cat,
      names_from = any_drug_exposure,
      values_from = str,
      values_fill = "0/0"
    ) |>
    rename(drug_exposed = `TRUE`, drug_unexposed = `FALSE`) |>
  select(days_since_exposure_cat, drug_unexposed, drug_exposed)
  p <- p + 1
}



listout[[1]] |>
  left_join(listout[[2]],
    by = "days_since_exposure_cat"
  ) |>
  mutate(across(everything(), \(x) if_else(is.na(x), "0/0", x))) |>
  kbl(
    booktabs = TRUE,
    col.names =
      c(
        "days since\ndrug exposure",
        rep(c("Echinocandin unexposed", "Echinocandin exposed"), 2)
      )
  ) |>
  add_header_above(c(
    " " = 1, "albicans" = 2, "glabrata" = 2
  )) |>
  kable_styling(latex_options = "scale_down")

```

```{r} 
#| label: tbl-acqres4
#| tbl-cap: "Number of isolates that have a log2 anidulafungin MIC fold change greater than 2, stratified by whether the patient has any echinocandin exposure at all, and time since echinocandin exposure. For those with no echinocandin exposure at all, the days since drug exposure variable is set to days since enrolment"
listout[[3]] |>
  left_join(listout[[4]],
    by = "days_since_exposure_cat"
  ) |>
  mutate(across(everything(), \(x) if_else(is.na(x), "0/0", x))) |>
  kbl(
    booktabs = TRUE,
    col.names =
      c(
        "days since\ndrug exposure",
        rep(c("Echinocandin unexposed", "Echinocandin exposed"), 2)
      )
  ) |>
  add_header_above(c(
    " " = 1, "dubliniensis" = 2, "parapsilosis" = 2
  )) |>
  kable_styling(latex_options = "scale_down")

```

```{r} 
#| label: tbl-acqres-bypt
#| tbl-cap: "Number of study participants with emergent resistance (defined by an increase in log2(mic fold change) of at least 2) to fluconazole stratified by azole exposure and species"


current_drug <- drug[[1]]
if (current_drug == "Azole") {
  current_mic_drug <- "fluconazole"
} else {
  current_mic_drug <- "anidulafungin"
}
    generate_mic_model_data(
      df_mics,
      df_drug,
      # spec = current_species,
      exposure_drug_class = current_drug,
      mic_drug = current_mic_drug,
      code_antifungal_exposure_as = "zero_at_first_exposure"
    )   |>
    group_by(psn, species) |>
    arrange(species, psn, swb_date) |>
    mutate(
      baseline_mic = first(mic),
    ) |> 
    filter(!(any_drug_exposure & days_since_first_drug_exposure < 0)) |>
    group_by(species, psn, any_drug_exposure) |>
    summarise(
      log2mic_inc= 
        if_else(any(log2(mic / baseline_mic) >= 2), ">=2","<2"),
      log2mic_inc = as.factor(log2mic_inc)) |>
    ungroup() |>
    mutate(any_drug_exposure = if_else(any_drug_exposure, "drug_exposed",
"drug_unexposed")) |>
  count(species, any_drug_exposure, log2mic_inc, .drop = "none") |>
    pivot_wider(id_cols = c(species, any_drug_exposure), 
    names_from = log2mic_inc,
    values_from = n)  |>
    nest(data = c(any_drug_exposure, `>=2`,`<2`)) |>
mutate(p = map(data, \(x) fisher.test(x[,2:3])$p.value)) |>
    unnest(cols = c(data, p)) |>
    rowwise() |>
    mutate(tot = `<2` + `>=2`,
           prop = `>=2` / tot,
           str = paste0(`>=2`, "/", tot, " (",round(prop*100,0),"%)")) |>
    select(species, any_drug_exposure, str, p) |>
    pivot_wider(id_cols = c(species, p),
    names_from = any_drug_exposure,
values_from = str) |>
    relocate(p, .after = everything()) |>
  kbl(
    booktabs = TRUE)
 

```

```{r} 
#| label: tbl-acqres-bypt2
#| tbl-cap: "Number of study participants with emergent resistance (defined by an increase in log2(mic fold change of at least 2) to anidulafungin stratified by echinocandin exposure and species"


current_drug <- drug[[2]]
if (current_drug == "Azole") {
  current_mic_drug <- "fluconazole"
} else {
  current_mic_drug <- "anidulafungin"
}
    generate_mic_model_data(
      df_mics,
      df_drug,
      # spec = current_species,
      exposure_drug_class = current_drug,
      mic_drug = current_mic_drug,
      code_antifungal_exposure_as = "zero_at_first_exposure"
    )   |>
    group_by(psn, species) |>
    arrange(species, psn, swb_date) |>
    mutate(
      baseline_mic = first(mic),
    ) |> 
    filter(!(any_drug_exposure & days_since_first_drug_exposure < 0)) |>
    group_by(species, psn, any_drug_exposure) |>
    summarise(
      log2mic_inc= 
        if_else(any(log2(mic / baseline_mic) >= 2), ">=2","<2"),
      log2mic_inc = as.factor(log2mic_inc)) |>
    ungroup() |>
    mutate(any_drug_exposure = if_else(any_drug_exposure, "drug_exposed",
"drug_unexposed")) |>
  count(species, any_drug_exposure, log2mic_inc, .drop = "none") |>
    pivot_wider(id_cols = c(species, any_drug_exposure), 
    names_from = log2mic_inc,
    values_from = n)  |>
    nest(data = c(any_drug_exposure, `>=2`,`<2`)) |>
mutate(p = map(data, \(x) fisher.test(x[,2:3])$p.value)) |>
    unnest(cols = c(data, p)) |>
    rowwise() |>
    mutate(tot = `<2` + `>=2`,
           prop = `>=2` / tot,
           str = paste0(`>=2`, "/", tot, " (",round(prop*100,0),"%)")) |>
    select(species, any_drug_exposure, str, p) |>
    pivot_wider(id_cols = c(species, p),
    names_from = any_drug_exposure,
values_from = str) |>
    relocate(p, .after = everything()) |>
  kbl(
    booktabs = TRUE)
 

```
\pagebreak 



## Modelling MIC with Gaussian process models.

@fig-micmods shows the parameter outputs from the Gaussian process models. The
key parameter is beta - this is the effect of antifungal exposure. There's no
great change from the previous models; a reduction in MIC to
fluconazole in glabrata and an increase in fluconazole MIC in albicans following
azole exposure, but not a strong effect - of the order of the standard deviation
of the MIC distribution. There is a clear within-participant temporal
correlation in MIC (@fig-miccov).

I think that the over all story from these models is that they do not identify a
strong effect of drug exposure in shifting the MIC distribution on the isolates.

```{r} 
#| label: fig-micmods
#| fig.cap: "Parameter values from MIC models with 95% CrI. (A) beta shows the effect of antimicrobial exposure on MIC [> 0, increases, < 0, decreases], alpha is the magnitude of the corelation between two samples within a participant, sigma_mic is the standard  deviation of the MIC distribution. (B) value of tau parameter (how long antimicrobial effect lasts for) (C) value of length_scale parameter that encodes the time that within-participant samples remain correlated for. In panel B and C the x axis variable is scaled time (helps to fit the model) - 1 unit there is 12.8 days"
#| fig.height:  8
#| fig.width: 8

fitted_mic_models <- 
  read_rds(here("data/fitted_stan_mic_gp_models.rda"))


bind_rows(
  lapply(
    fitted_mic_models,
    function(x) {
      bind_cols(
        species = x$species,
        drug = x$drug,
        bayesplot::mcmc_intervals_data(x$model_fit$draws(),
          regex_pars = "sigma|beta|tau|length|alpha"
        )
      )
    }
  )
) -> df

p_a <-
df |>
  filter(!grepl("lp_|tau|length", parameter)) |>
  ggplot(aes(parameter, m, ymin = ll, ymax = hh, color = species)) +
  geom_point(position = position_dodge(0.5)) +
  geom_errorbar(width = 0, position = position_dodge(0.5)) +
  facet_wrap(~drug) +
  coord_flip() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(y = "parameter value") +
geom_hline(aes(yintercept = 0), linetype = "dashed") 

p_b <-
df |>
  filter(grepl("tau", parameter)) |>
  ggplot(aes(parameter, m, ymin = ll, ymax = hh, color = species)) +
  geom_point(position = position_dodge(0.3)) +
  geom_errorbar(width = 0, position = position_dodge(0.3)) +
  facet_wrap(~drug) +
  coord_flip()+
  theme_bw() +
  scale_color_viridis_d() +
geom_hline(aes(yintercept = 0), linetype = "dashed") 

p_c <-
df |>
  filter(grepl("length", parameter)) |>
  ggplot(aes(parameter, m, ymin = ll, ymax = hh, color = species)) +
  geom_point(position = position_dodge(0.3)) +
  geom_errorbar(width = 0, position = position_dodge(0.3)) +
  facet_wrap(~drug) +
  coord_flip() +
  theme_bw() +
  scale_color_viridis_d() +
geom_hline(aes(yintercept = 0), linetype = "dashed") 

p_a / p_b / p_c + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")

```

```{r time-corr}
#| label: fig-miccov
#| fig.height: 5
#| fig.width: 8
#| fig-cap: "Posterior of the fitted covariance function between  two samples from the same participant. Each line is one draw from the posterior. Time here is scaled so one unit it 12.8 days"
#| dev: 'jpeg'

bind_rows(
  lapply(
    fitted_mic_models,
    function(x) {
      bind_cols(
        species = x$species,
        drug = x$drug,
        x$model_fit$draws(format = "df")
        )
    }
    )
) -> df

cross_join(
  tibble(x = seq(0, 10, by = 0.1)),
  df |>
    select(c(species, drug, length_scale, alpha, .draw))
) |>
  mutate(f = alpha^2 * exp(-x^2 / (2 * length_scale^2))) |>
  ggplot(aes(x, f, group = .draw, color = species)) +
  geom_line(alpha = 0.01) +
  facet_grid(drug ~ species) +
  theme_bw() +
  scale_color_viridis_d() +
  theme(legend.position = "none") +
  labs(x = "time")


```

# Conclusions

I think that main message from these models is that there is not a strong
population shift of MIC following drug exposure. It's possible that there are
some interesting effects of azole exposure on fluconazole MIC but these could
very likely be affected by confounding and they effects are not particularly
strong when compared to the SD of the MIC distribution. But drug exposure is not
driving a large population shift in MIC.

However, the simpler analysis does suggest that (for azole in particular) drug
exposure is associated with sporadic emergence of increase in MIC by our
arbitrary threshold (at least log~2~ [fold MIC change] of 2). This is I guess
consistent with stochastic emergence of resistance with sequential mutation or
acquisition of genes - for which the sequencing may be a helpful next step.




